Minimax-optimal Laplacian Eigenmaps regression over Sobolev Spaces with Neighborhood Graphs

### 2021-08-23 manuscripts/laplacian_eigenmaps/laplacian_eigenmaps_regression.tex
Updated preliminaries section of Laplacian Eigenmaps manuscript.

\paragraph{Estimation.} When the kernel $\eta$, graph radius $\varepsilon$, and number of eigenvectors $K$ are chosen appropriately, we show in Theorem~\ref{thm:laplacian_eigenmaps_estimation_fo} that the estimator $\wh{f}$ achieves the minimax rate over $H^1(\mc{X};M)$, when the error is measured in squared $L^2(P_n)$ norm.
\begin{enumerate}[label=(K\arabic*)]
	\setcounter{enumi}{0}
	\item
	\label{asmp:kernel_flat_euclidean}
	The kernel function $\eta$ is a nonincreasing function supported on $[0,1]$. Its restriction to $[0,1]$ is Lipschitz, and $\eta(1) > 0$. Additionally, it is normalized so that
	\begin{equation*}
	\int_{\Rd} \eta(\|z\|) \,dz = 1.
	\end{equation*}
	and we assume \smash{$\sigma_{\eta} := \frac{1}{d}\int_{\Rd} \|x\|^2 \eta(\|x\|) \,dx < \infty$}.
\end{enumerate}
\begin{enumerate}[label=(P\arabic*)]
	\setcounter{enumi}{0}
	\item 
	\label{asmp:parameters_estimation_fo} 
	For constants $c_0$ and $C_0$, the graph radius $\varepsilon$ and the number of eigenvectors $K$ satisfy the following inequalities:
	\begin{equation}\\
	\label{eqn:radius_fo} 
	C_0\biggl(\frac{\log n}{n}\biggr)^{1/d} \leq \varepsilon \leq c_0\min\{1,K^{-1/d}\},
	\end{equation}
	and 
	\begin{equation}
	\label{eqn:eigenvector_estimation_fo} 
	K = \min\Bigl\{\floor{(M^2n)^{d/(2 + d)}} \vee 1, n\Bigr\}.
	\end{equation}
\end{enumerate}
We comment on these assumptions after stating our first main theorem, regarding the estimation error of Laplacian eigenmaps.
\begin{theorem}
	\label{thm:laplacian_eigenmaps_estimation_fo}
	Suppose Model~\ref{def:model_flat_euclidean}, and additionally $f_0 \in H^1(\mc{X},M)$. There are constants $c,C$ and $N$ (not depending on $f_0$, $M$ or $n$), such that the following statement holds for all $n \geq N$ and any $\delta \in (0,1)$: if the Laplacian eigenmaps estimator $\wh{f}$ is computed with a kernel $\eta$ satisfying~\ref{asmp:kernel_flat_euclidean}, and parameters $\varepsilon$ and $K$ satisfying~\ref{asmp:parameters_estimation_fo}, then
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_estimation_fo}
	\|\wh{f} - f_0\|_n^2 \leq C\Bigl(\frac{1}{\delta}M^2(M^2n)^{-2/(2 + d)} \wedge 1\Bigr) \vee \frac{1}{n},
	\end{equation}
	with probability at least $1 - \delta - Cn\exp(-cn\varepsilon^d) - \exp(-K)$.
\end{theorem}
From~\eqref{eqn:laplacian_eigenmaps_estimation_fo} it follows immediately that with constant probability $\|\wh{f} - f_0\|_n^2 \lesssim M^2(M^2n)^{-2/(2 + d)}$ whenever $n^{-1/2} \lesssim M \lesssim n^{1/d}$.

@@

	\item The upper bound given in equation~\eqref{eqn:laplacian_eigenmaps_estimation_fo} holds with probability $1 - \delta - Cn\exp(-cn\varepsilon^d)$. Under the stronger assumption that $f_0 \in C^1(\mc{X};M)$ we can replace the factor of $\delta$ by the sharper $\delta^2/n$, which is less than $\delta$ because $\delta \in (0,1)$. Then a routine calculation, which we carry out in Appendix~\ref{subsec:upper_bound_l2pn_risk}, shows that the expected $L^2(P_n)$ loss is on the same order as~\eqref{eqn:laplacian_eigenmaps_estimation_fo}, matching the minimax rate.
\end{itemize}

@@

We now outline the high-level strategy we follow when proving each of Theorems~\ref{thm:laplacian_eigenmaps_estimation_fo}-\ref{thm:laplacian_eigenmaps_testing_ho_suboptimal}. We analyze the estimation error of $\wh{f}$, and the testing error of $\wh{\varphi}$, by first conditioning on the design points $X_1,\ldots,X_n$ and deriving \emph{design-dependent} bias and variance terms. For estimation, we have that with probability at least $1 - \exp(-K)$,
\begin{equation}
\label{eqn:estimation_biasvariance}
\|\wh{f} - f_0\|_n^2 \leq \underbrace{\frac{\dotp{L_{n,\varepsilon}^s f_0}{f_0}_n}{\lambda_{K}^s}}_{\textrm{bias}} + \underbrace{\frac{5K}{n} \vphantom{\frac{\dotp{L^s f_0}{f_0}_n}{\lambda_{K}^s}}}_{\textrm{variance}}.
\end{equation}
For testing, we have that $\varphi$ (which is a level-$a$ test by construction) also has small Type II Error, $\Ebb_{f_0}[1 - \phi] \leq b/2$, if 
\begin{equation}
\label{eqn:testing_biasvariance}
\|f_0\|_n^2 \geq  \underbrace{\frac{\dotp{L_{n,\varepsilon}^s f_0}{f_0}_n}{\lambda_{K}^s}}_{\textrm{bias}} + \underbrace{32\frac{\sqrt{2K}}{n}\biggl[\sqrt{\frac{1}{a}} + \frac{1}{b}\biggr]}_{\textrm{variance}}.
\end{equation}
The quadratic form $\dotp{L_{n,\varepsilon}^s f_0}{f_0}_n$, eigenvalue $\lambda_{K}$, and empirical squared norm $\|f_0\|_n^2$ are each random variables that depend on the random design points $X_1,\ldots,X_n$. We proceed to establish suitable upper and lower bounds on these quantities.

@@

\begin{enumerate}[label=(P\arabic*)]
	\setcounter{enumi}{4}
	\item 
	\label{asmp:kernel_manifold}
	The kernel function $\eta$ is a nonincreasing function supported on a subset of $[0,1]$. Its restriction to $[0,1]$ is Lipschitz, and $\eta(1/2) > 0$. Additionally, it is normalized so that
	\begin{equation*}
	\int_{\Reals^m} \eta(\|z\|) \,dz = 1,
	\end{equation*}
	and we assume \smash{$\int_{\Reals^m} \|x\|^2 \eta(\|x\|) \,dx < \infty$}.
	\item 
	\label{asmp:parameters_estimation_manifold}
	For a constant $C_0$, the graph radius $\varepsilon$ and number of eigenvectors $K$ satisfy
	\begin{equation}
	\label{eqn:radius_estimation_manifold}
	C_0\max\biggl\{\biggl(\frac{\log}{n}\biggr)^{1/m}, n^{-1/(2(s - 1) + m)}\biggr\} \leq \varepsilon \leq \min\{i_0, K^{-1/m}\},
	\end{equation}
	where we recall that $i_0$ is a lower bound on the injectivity radius of $\mc{X}$. Additionally,
	\begin{equation*}
	K = \min\Bigl\{\floor{(M^2n)^{m/(2s + m)}} \wedge 1,n \Bigr\}.
	\end{equation*}
\end{enumerate}

>>>>

\paragraph{Estimation.} When the kernel $\eta$, graph radius $\varepsilon$, and number of eigenvectors $K$ are chosen appropriately, we show in Theorem~\ref{thm:laplacian_eigenmaps_estimation_fo} that the estimator $\wh{f}$ achieves the minimax rate over $H^1(\mc{X};M)$, when the error is measured in squared $L^2(P_n)$ norm.
\begin{enumerate}[label=(K\arabic*)]
	\setcounter{enumi}{0}
	\item
	\label{asmp:kernel_flat_euclidean}
	The kernel function $\eta$ is a nonincreasing function supported on $[0,1]$. Its restriction to $[0,1]$ is Lipschitz, and $\eta(1) > 0$. Additionally, it is normalized so that
	\begin{equation*}
	\int_{\Rd} \eta(\|z\|) \,dz = 1.
	\end{equation*}
	and we assume \smash{$\sigma_{\eta} := \frac{1}{d}\int_{\Rd} \|x\|^2 \eta(\|x\|) \,dx < \infty$}.
\end{enumerate}
\begin{enumerate}[label=(P\arabic*)]
	\setcounter{enumi}{0}
	\item 
	\label{asmp:parameters_estimation_fo} 
	For constants $c_0$ and $C_0$, the graph radius $\varepsilon$ and the number of eigenvectors $K$ satisfy the following inequalities:
	\begin{equation}\\
	\label{eqn:radius_fo} 
	C_0\biggl(\frac{\log n}{n}\biggr)^{1/d} \leq \varepsilon \leq c_0\min\{1,K^{-1/d}\},
	\end{equation}
	and 
	\begin{equation}
	\label{eqn:eigenvector_estimation_fo} 
	K = \min\Bigl\{\floor{(M^2n)^{d/(2 + d)}} \vee 1, n\Bigr\}.
	\end{equation}
\end{enumerate}
We comment on these assumptions after stating our first main theorem, regarding the estimation error of Laplacian eigenmaps.
\begin{theorem}
	\label{thm:laplacian_eigenmaps_estimation_fo}
	Suppose Model~\ref{def:model_flat_euclidean}, and additionally $f_0 \in H^1(\mc{X},M)$. There are constants $c,C$ and $N$ (not depending on $f_0$, $M$ or $n$), such that the following statement holds for all $n \geq N$ and any $\delta \in (0,1)$: if the Laplacian eigenmaps estimator $\wh{f}$ is computed with a kernel $\eta$ satisfying~\ref{asmp:kernel_flat_euclidean}, and parameters $\varepsilon$ and $K$ satisfying~\ref{asmp:parameters_estimation_fo}, then
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_estimation_fo}
	\|\wh{f} - f_0\|_n^2 \leq C\Bigl(\frac{1}{\delta}M^2(M^2n)^{-2/(2 + d)} \wedge 1\Bigr) \vee \frac{1}{n},
	\end{equation}
	with probability at least $1 - \delta - Cn\exp(-cn\varepsilon^d) - \exp(-K)$.
\end{theorem}
From~\eqref{eqn:laplacian_eigenmaps_estimation_fo} it follows immediately that with constant probability $\|\wh{f} - f_0\|_n^2 \lesssim M^2(M^2n)^{-2/(2 + d)}$ whenever $n^{-1/2} \lesssim M \lesssim n^{1/d}$, matching the standard minimax estimation rate over Sobolev classes.

@@

	\item The upper bound given in equation~\eqref{eqn:laplacian_eigenmaps_estimation_fo} holds with probability $1 - \delta - Cn\exp(-cn\varepsilon^d)$. Under the stronger assumption that $f_0 \in C^1(\mc{X};M)$ we can replace the factor of $\delta$ by the sharper $\delta^2/n$, which is less than $\delta$ because $\delta \in (0,1)$. Then a routine calculation shows that the expected $L^2(P_n)$ loss is on the same order as~\eqref{eqn:laplacian_eigenmaps_estimation_fo}, matching the minimax rate.
\end{itemize}

@@

We now outline the high-level strategy we follow when proving each of Theorems~\ref{thm:laplacian_eigenmaps_estimation_fo}-\ref{thm:laplacian_eigenmaps_testing_ho_suboptimal}. We analyze the estimation error of $\wh{f}$, and the testing error of $\wh{\varphi}$, by first conditioning on the design points $X_1,\ldots,X_n$ and deriving \emph{design-dependent} bias and variance terms. For estimation, we have that with probability at least $1 - \exp(-K)$,
\begin{equation}
\label{eqn:estimation_biasvariance}
\|\wh{f} - f_0\|_n^2 \leq \underbrace{\frac{\dotp{L_{n,\varepsilon}^s f_0}{f_0}_n}{\lambda_{K}^s}}_{\textrm{bias}} + \underbrace{\frac{5K}{n} \vphantom{\frac{\dotp{L^s f_0}{f_0}_n}{\lambda_{K}^s}}}_{\textrm{variance}}.
\end{equation}
For testing, we have that $\varphi$ (which is a level-$a$ test by construction) also has small Type II Error, $\Ebb_{f_0}[1 - \phi] \leq b/2$, if 
\begin{equation}
\label{eqn:testing_biasvariance}
\|f_0\|_n^2 \geq  \underbrace{\frac{\dotp{L_{n,\varepsilon}^s f_0}{f_0}_n}{\lambda_{K}^s}}_{\textrm{bias}} + \underbrace{32\frac{\sqrt{2K}}{n}\biggl[\sqrt{\frac{1}{a}} + \frac{1}{b}\biggr]}_{\textrm{variance}}.
\end{equation}
The quadratic form $\dotp{L_{n,\varepsilon}^s f_0}{f_0}_n$, eigenvalue $\lambda_{K}$, and empirical squared norm $\|f_0\|_n^2$ are each random variables that depend on the random design points $X_1,\ldots,X_n$. We proceed to establish suitable upper and lower bounds on these quantities. 
% (AG 8/19/21): Need to include some discussion of how the analysis contrasts with that of the classical spectral projection estimator and test.

@@

\begin{enumerate}[label=(P\arabic*)]
	\setcounter{enumi}{4}
	\item 
	\label{asmp:kernel_manifold}
	The kernel function $\eta$ is a nonincreasing function supported on a subset of $[0,1]$. Its restriction to $[0,1]$ is Lipschitz, and $\eta(1/2) > 0$. Additionally, it is normalized so that
	\begin{equation*}
	\int_{\Reals^m} \eta(\|z\|) \,dz = 1,
	\end{equation*}
	and we assume \smash{$\int_{\Reals^m} \|x\|^2 \eta(\|x\|) \,dx < \infty$}.
	\item 
	\label{asmp:parameters_estimation_manifold}
	For a constant $C_0$, the graph radius $\varepsilon$ and number of eigenvectors $K$ satisfy
	\begin{equation}
	\label{eqn:radius_estimation_manifold}
	C_0\max\biggl\{\biggl(\frac{\log}{n}\biggr)^{1/m}, n^{-1/(2(s - 1) + m)}\biggr\} \leq \varepsilon \leq \min\{\mathrm{inj}(\mc{X}), K^{-1/m}\},
	\end{equation}
	where we recall that $\mathrm{inj}(\mc{X})$ is a lower bound on the injectivity radius of $\mc{X}$. Additionally,
	\begin{equation*}
	K = \min\Bigl\{\floor{(M^2n)^{m/(2s + m)}} \wedge 1,n \Bigr\}.
	\end{equation*}
\end{enumerate}
